---
title: "fMRI_VAR"
output:
  pdf_document: default
  html_document: default
date: "2024-05-12"
---

```{r setup, include=FALSE}
#### 1. Import libraries ####
library(reshape2)
library(gdata)
library(DescTools)
library(R.matlab)
library(data.table)
library(zoo)
library(MTS)

#### 2. Paths ####
source("~/Documents/Cornell/Projects/HCP/Code/helper_function.R")
load("~/Dropbox/HCP/elvisha/SC/sc_sdstream_fs86_997subj.RData")
out_path <- "~/Documents/Cornell/Courses/TS and SP/Project/"
ts_input <- "~/Dropbox/HCP/elvisha/timeseries/RData/"
regions <- as.vector(read.table("~/Dropbox/HCP/regions.txt")$V1)
hcp.dir <- "~/Dropbox/HCP/elvisha/"
```


## Background

VAR analysis of fMRI data on healthy participants. Very briefly, p = 86,
T = 1200, TR= 720 ms for each subject.  We focus on p = 6,
T = 1200 for each subject. 


## Data cleaning

```{r}
#### 3. Get subject id #### 
fc_subjects <- sub("^(.*?)_.*$", "\\1", 
                   list.files(path = ts_input)) 
sc_subjects <- as.character(dat.SC$subject)
subjects <- intersect(fc_subjects, sc_subjects)
length(subjects)
```

```{r}
#### 4. Extract FC and SC #### 
n = length(subjects)
r = length(regions) 
tp = 1200
TS <- array(0, c(n,tp,r))
SC <- array(0, c(n,r,r))

for (i in 1:n) {
  id <- subjects[i]
  # print(id)
  
  # Generate FC
  load(paste0(ts_input, id, "_TS.RData"))
  # 4 replicates 4 x 1200 time points = 4800
  ts_list <- list()
  for (j in 1:4) {
    reps <- nrow(timeseries)/4
    ts <- timeseries[((j-1)*reps + 1):(j*reps), 1:r] 
    if (all(is.na(ts))) {
      next
    } else {
      # pearson cor
      ts_list[[j]] <- ts
    }
  }
  
  # remove null replicate
  ts_list[sapply(ts_list, is.null)] <- NULL


  if (length(ts_list) == 4) {
    # average 4 ts 
    TS[i,,] <- as.matrix(ts_list[[1]] + ts_list[[2]] + 
                           ts_list[[3]] +ts_list[[4]]) / 4 
  } else {
    next
  }
  
  # Get SC sift2volnorm version
  SC[i,,] <- as.array(dat.SC$sift2volnorm[which(dat.SC$subject==as.numeric(id)), ,])
}

dim(TS)

# Combine all regions
regions <- data.table(r_num = 1:r, r_name = regions)
```

## Focus on 6 subcortical regions
```{r}
subcort <- c("Thalamus-Proper", "Hippocampus", "Amygdala")
subcort_indx <- which(grepl(paste(subcort, collapse="|"), regions$r_name))
subcort_r <- regions[subcort_indx, r_name]
TS_subcort <- TS[, , subcort_indx]
SC_subcort <- SC[, , subcort_indx]
```

## Plot TS of 1 subject

All times series
```{r}
i = 1
x <- data.frame(TS_subcort[i, ,])
dim(x)
names(x) <- c("L-Thalamus", "L-Hippo", "L-Amyg",
              "R-Thalamus", "R-Hippo", "R-Amyg")
head(x)

# normalize the data
x <- scale(x)

par(mfrow = c(1,2))
matplot(x, type = 'l', xlab="Time", ylab="")
# ts.plot(x)
plot.ts(x, main="")
par(mfrow = c(1,1))
```


## ACF and PACF
L- and R-Thalamus: ACF shows exponential decay, PACF shows significant up to lag 7
L-Hippo: ACF also shows exponential decay, slightly significant at the end, PACF shows significant at 1, 2, 27, 29
R-Hippo: ACF shows exponential decay, PACF shows significant up to lag 3
L-Amyg: ACF shows exponential decay, PACF shows significant up to lag 7
R-Amyg: ACF shows exponential decay, PACF shows significant up to lag 3

```{r}
for (j in 1:6) {
  par(mfrow = c(1,2))
  acf(x[, j])
  pacf(x[, j])
  par(mfrow = c(1,1))
}
```
## Cross correlation

Separate into each hemisphere

```{r}
left_x <- x[, 1:3]

matplot(left_x, type = 'l', xlab="Time", ylab="")
ts.plot(left_x)
names(left_x) <- gsub("Left-", "", names(left_x))
plot.ts(left_x, main="")

acf(left_x)              # cross-corrleation function 2x2
acf(left_x[, 1:2])              # cross-corrleation function 2x2
acf(left_x[, 2:3])              # cross-corrleation function 2x2
acf(left_x[, c(1,3)])              # cross-corrleation function 2x2

```

```{r}
right_x <- x[, 4:6]

matplot(right_x, type = 'l', xlab="Time", ylab="")
ts.plot(right_x)
names(right_x) <- gsub("Right-", "", names(right_x))
plot.ts(right_x, main="")

acf(right_x)              # cross-corrleation function 2x2
acf(right_x[, 1:2])              # cross-corrleation function 2x2
acf(right_x[, 2:3])              # cross-corrleation function 2x2
acf(right_x[, c(1,3)])              # cross-corrleation function 2x2
```

Same region in each hemisphere
```{r}
thal_x <- x[, c(1,4)]
acf(thal_x)              # cross-corrleation function 2x2

hippo_x <- x[, c(2,5)]
acf(hippo_x)              # cross-corrleation function 2x2

amyg_x <- x[, c(3,6)]
acf(amyg_x)              # cross-corrleation function 2x2

```
## VAR model
Ljung-Box test 
Note that it is applied to the residuals of a fitted ARIMA model, not the original series, and in such applications the hypothesis actually being tested is that the residuals from the ARIMA model have no autocorrelation.
Use Augmented Dickey-Fuller Test
```{r}
# mq(x, lag = 5)    
library(fpp)
for (j in 1:6) {
  print(adf.test(x[,j]))
}
```

VAR(p) model
```{r}
y <- as.ts(x)  

#acf(y)

fit.VARp <- VARorder(y, maxp = 10, output = T)

plot(1:11, fit.VARp$aic, type = 'l')
points(which.min(fit.VARp$aic)-1, fit.VARp$aic[which.min(fit.VARp$aic)], pch = 20, col = 2, cex = 3)
which.min(fit.VARp$aic) # (subtract 1 for lag order since lag 0 is included) 
```

```{r}
# Fit a VAR(2) model 
VAR(y, p = 2, output = T, include.mean = F, fixed = NULL)
fit.VAR2 = VAR(y, p = 2, output = T, include.mean = F, fixed = NULL)
names(fit.VAR2)
?VAR
fit.VAR2$coef
fit.VAR2$Phi
fit.VAR2$Ph0
fit.VAR2$Sigma

# se's for regression parameters:
fit.VAR2$secoef

# t-ratios for regression parameters:
t = fit.VAR2$coef / fit.VAR2$secoef

# Estimated residual series
a.hat <- fit.VAR2$residuals  
head(a.hat)
nrow(y)
nrow(a.hat) # removed first 2 rows since we fit VAR(2)

```

$$
\begin{pmatrix}
y_{1,t} \\
y_{2,t} \\
y_{3,t} \\
y_{4,t} \\
y_{5,t} \\
y_{6,t} 
\end{pmatrix}
=
\begin{pmatrix}
0.1168 & 0.0525 & 0.0501 & 0.19584 & 0.0307 & -0.0150 \\
-0.0178 & 0.2119 & 0.0704 & 0.00892 & 0.1588 & -0.0124 \\
-0.0582 & 0.0651 & 0.0939 & -0.03514 & 0.0259 & 0.0756 \\
0.1457 & 0.0435 & 0.0257 & 0.23802 & 0.0388 & -0.0541 \\
-0.0316 & 0.1655 & 0.0383 & 0.03441 & 0.2077 & -0.0368 \\
0.0181 & 0.0560 & 0.1099 & -0.04380 & 0.0850 & 0.1249
\end{pmatrix}
\begin{pmatrix}
y_{1,t-1} \\
y_{2,t-1} \\
y_{3,t-1} \\
y_{4,t-1} \\
y_{5,t-1} \\
y_{6,t-1} 
\end{pmatrix} +

\\ 
\begin{pmatrix}
  0.0932 & -0.02069 & -0.0381 & 0.11332 & 0.02599 & -0.00169 \\
 -0.0476 & 0.17639 & 0.0628 & -0.01739 & 0.02958 & -0.01498 \\
 -0.0393 & 0.03743 & 0.0905 & -0.01315 & -0.00625 & 0.06363 \\
  0.0621 & 0.00252 & -0.0206 & 0.10051 & 0.02675 & -0.01700 \\
  0.0535 & 0.13834 & 0.0154 & -0.04180 & 0.10613 & 0.01436 \\
  0.0100 & -0.01156 & 0.0901 & -0.00904 & -0.03243 & 0.09979
\end{pmatrix}
\begin{pmatrix}
y_{1,t-2} \\
y_{2,t-2} \\
y_{3,t-2} \\
y_{4,t-2} \\
y_{5,t-2} \\
y_{6,t-2} 
\end{pmatrix} +
\begin{pmatrix}
a_{1,t} \\
a_{2,t} \\
a_{3,t} \\
a_{4,t} \\
a_{5,t} \\
a_{6,t} 
\end{pmatrix} 
\\
\text{ where } Cov(a_t) = \Sigma = \begin{pmatrix}
  0.8476 & 0.0027 & 0.0446 & 0.2093 & 0.0523 & -0.0242 \\
  0.0027 & 0.7865 & 0.0499 & 0.0208 & 0.2036 & 0.0513 \\
  0.0446 & 0.0499 & 0.9295 & -0.0183 & 0.0551 & 0.1239 \\
  0.2093 & 0.0208 & -0.0183 & 0.8150 & 0.0764 & -0.0170 \\
  0.0523 & 0.2036 & 0.0551 & 0.0764 & 0.7849 & 0.0965 \\
 -0.0242 & 0.0513 & 0.1239 & -0.0170 & 0.0965 & 0.9070 \\
\end{pmatrix}



$$


# Diagonostic checking

```{r}
# Diagnostic checking
par(mfrow = c(2,2))
ccf(a.hat[,1],a.hat[,2])
matplot(a.hat, type = 'l')
plot(a.hat[,1], type = 'l') ; abline(h = 0)
plot(a.hat[,2], type = 'l') ; abline(h = 0)
par(mfrow = c(1,1))
#

plot(a.hat[,1],a.hat[,2])
plot(a.hat[,1],a.hat[,4])


par(mfrow = c(2,2))
hist(a.hat[,1])
hist(a.hat[,2])
hist(a.hat[,3])

hist(a.hat[,4])
hist(a.hat[,5])
hist(a.hat[,6])


qqnorm(a.hat[,1]) ; qqline(a.hat[,1])
qqnorm(a.hat[,2]) ; qqline(a.hat[,2])
qqnorm(a.hat[,3]) ; qqline(a.hat[,3])

qqnorm(a.hat[,4]) ; qqline(a.hat[,4])
qqnorm(a.hat[,5]) ; qqline(a.hat[,5])
qqnorm(a.hat[,6]) ; qqline(a.hat[,6])

par(mfrow = c(1,1))

#
acf(a.hat)

mq(a.hat, lag = 5)    

grangertest(a.hat[,1], a.hat[,4], order = 2)
grangertest(a.hat[,4], a.hat[,1], order = 2)

grangertest(a.hat[,2], a.hat[,5], order = 2)
grangertest(a.hat[,5], a.hat[,2], order = 2)

grangertest(a.hat[,3], a.hat[,6], order = 2)
grangertest(a.hat[,6], a.hat[,3], order = 2)

```

```{r}
# Multivariate Ljung-Box Q(m) statistics w and w/o adjustment
# g <- p * k^2 
g <- 2*2^2 ; g  # number of regression parameters (ignoring intercept)
mq(a.hat,10)
mq(a.hat,10, adj = g)
mq(a.hat,15)
mq(a.hat,15, adj = g)

# MTSdiag(fit.VAR2, gof = 24, adj = 0, level = F)

# acf(a.hat^2)
MarchTest(a.hat, lag = 10)


# Refine a fitted VAR model by removing simultaneously insignificant parameters
refVAR(fit.VAR2, fixed = NULL, thres = 2)
```

## Fit model for all participants 

```{r}
ar1 <- array(0, c(n,6,6))
ar2 <- array(0, c(n,6,6))

for (i in 1:49) {
  x <- data.frame(TS_subcort[i, ,])
  names(x) <- c("L-Thalamus", "L-Hippo", "L-Amyg",
                "R-Thalamus", "R-Hippo", "R-Amyg")
  
  # normalize the data
  x <- scale(x)
  
  y <- as.ts(x)
  fit.VAR2 = MTS::VAR(y, p = 2, include.mean = F, fixed = NULL, output=F)
  refit.VAR2 = MTS::refVAR(fit.VAR2, fixed = NULL, thres = 2)
  ar1[i, , ] <- t(refit.VAR2$coef[1:6,])
  ar2[i, , ] <- t(refit.VAR2$coef[7:12,])
  
}
```
# Correlation of ar1 with SC
```{r}

cor(scale((SC_subcort[, 4,1])), ar1[, 4, 1])
cor(scale((SC_subcort[, 5,2])), ar1[, 5, 2])
cor(scale((SC_subcort[, 6,3])), ar1[, 6, 3])

par(mfrow = c(3,1))
plot(scale((SC_subcort[, 4,1])), ar1[, 4, 1],
     ylab="Thalamus",
     xlab="", xlim=c(-3,3), ylim=c(0,0.5))
plot(scale((SC_subcort[, 5,2])), ar1[, 5, 2],
     ylab="Hippocampus",
     xlab="", xlim=c(-3,3), ylim=c(0,0.5))
plot(scale((SC_subcort[, 6,3])), ar1[, 6, 3],
     ylab="Amygdala",
     xlab="Scaled structural connectivity",
     xlim=c(-3,3), ylim=c(0,0.5))
par(mfrow = c(1,1))

```

# Reverse
```{r}

cor(scale(log(SC_subcort[, 4,1])), ar1[, 1, 4])
cor(scale(log(SC_subcort[, 5,2])), ar1[, 2, 5])
cor(scale(log(SC_subcort[, 6,3])), ar1[, 3, 6])

par(mfrow = c(3,1))
plot(scale((SC_subcort[, 4,1])), ar1[, 1, 4],
     ylab="Thalamus",
     xlab="", xlim=c(-3,3), ylim=c(0,0.5))
plot(scale((SC_subcort[, 5,2])), ar1[, 2, 5],
     ylab="Hippocampus",
     xlab="", xlim=c(-3,3), ylim=c(0,0.5))
plot(scale((SC_subcort[, 6,3])), ar1[, 3, 6],
     ylab="Amygdala",
     xlab="Scaled log structural connectivity",
     xlim=c(-3,3), ylim=c(0,0.5))
par(mfrow = c(1,1))

```

# Correlation of ar2 with SC
```{r}

cor(scale((SC_subcort[, 4,1])), ar2[, 4, 1])
cor(scale((SC_subcort[, 5,2])), ar2[, 5, 2])
cor(scale((SC_subcort[, 6,3])), ar2[, 6, 3])

par(mfrow = c(3,1))
plot(scale((SC_subcort[, 4,1])), ar2[, 4, 1],
     ylab="Thalamus",
     xlab="", xlim=c(-3,3), ylim=c(0,0.5))
plot(scale((SC_subcort[, 5,2])), ar2[, 5, 2],
     ylab="Hippocampus",
     xlab="", xlim=c(-3,3), ylim=c(0,0.5))
plot(scale((SC_subcort[, 6,3])), ar2[, 6, 3],
     ylab="Amygdala",
     xlab="Scaled structural connectivity",
     xlim=c(-3,3), ylim=c(0,0.5))
par(mfrow = c(1,1))

```

